\documentclass{amsart}

\usepackage{wasysym}
%\usepackage{ifpdf}
%\usepackage{mathpartir}
\usepackage{amssymb}
%\usepackage[all]{xypic}
%\xyoption{2cell}
%\xyoption{curve}
%\UseTwocells
\input{diagxy}

% \usepackage{makeindex}

\include{thesis-macros}
\include{thesis-style}

\begin{document}

\section{Bibliography}

Citation styles\ldots\ I don't like \textbf{[17]}---it makes them a complete black box, useless unless one flips constantly back and forth.  I quite like \textbf{[Batanin 2009]}, but it takes up a lot of space, so obliges one to cite less liberally; also it's perhaps a little eccentric.  Overall, \textbf{[Bat09]}\ probably a decent compromise---it's readable for someone who knows the literature a little, and to someone who doesn't, nothing short of the full title is much more useful anyway.

Author naming: full name, or surname only?  I prefer full names; again, since bulkier, this means using them a bit sparinglier than with surnames alone, but since I'm using mostly-readable citation style, I don't have to use names too often, so this seems OK.

\section{Sectioning}  numbering: follow Cisinski[?] and number just by paragraph within section?  Even within chapter??  Ask Michael \& Steve about this first...  but I think something like this is nice; it makes the often artificial distinction between numbered definitions/propositions and open discussion much less consequential.

OK!  Here's what I think I want (5.vii):

chapters, section: as usual.

Subsections --- use just for their headings, \emph{not} numbered.  (display in toc?  undecided!)

Theorems, definitions, pars: \emph{all} numbered consecutively within section.
Deliniated paras: numbered/named in bold, with no indent; ordinary paras: small indent.  As in: \\

\noindent \textbf{3.2\ Id-Types are awesome.}\ \ I think that $\Id$-types are awesome, and you should think so too!

The most fundamentally compelling reason for this is that\ldots \\
  
\noindent \textbf{3.3}\ \ However, if we look at some of the alternative eliminators for $\Id$-types, \ldots \\

Hmmm\ldots but what should (sub)section headings look like?  Bold and/or small caps?  Centred and/or left-aligned?  Probably: both centred, section in bold small caps, subsections (if used) in just bold?  (So: all amsbook default, except making sections bold (and maybe also giving them more space?), for better prominence.)

Indentation: the amsbook paragraph indent is rather big!  Make it like in amsart instead?  YES!

Also note: text beginning directly after a (sub)section heading thus needs a \texttt{\\newline} and a \texttt{\\noindent}.  (Maybe just put a \texttt{\\newline} in the definiton of subsection somehow?)  CHECKLIST at end: make sure of this!

Running page headers and references: if numbering paras (etc.) within sections,  reference that as [2.3.11] (and show it in running header for easy navigation?)

\para Figures should also, surely, then be numbered in sequence with everything else.  Is it possible to make \LaTeX{} do this?

\section{Notations}
\para The classifying $\CwA_\strat$ of a theory: $\cl_\strat(\TT)$?  Problem: $\cl$ looks like ``closure'' not classifying!  $\C(\TT)$ or $\CC(\TT)$ is more usual, but not brilliant by any means\ldots\ and gets worse as we decorate it $\C_\strat$, $\C_\omega$, $\C^-_\omega$, etc.\ldots

At least, identifying a theory transparently with its classifying (strat) CwA lets us avoid using $\cl$ except for as $\cl_\omega$.  

\para The boundary of a pasting diagram??  Batanin and others use $\delta \pi$ to mean the source/target of $\pi$ (since they're equal).  I DON'T LIKE THIS: several reasons!  Firstly, it's misleading: it's not the boundary in the usual well-established sense of $\delta$ for discs; that's the pushout of the source and target along their common boundary.  Secondly, it thus leaves us \emph{without} a natural notation for that boundary, even though it's something we use all the time!  Thirdly (this is minor and debatable) it's often helpful to distinguish $s\pi$ and $t\pi$ when working with them, since they have different natural embeddings into $\pi$, etc.

I certainly want to avoid supporting this usage: I'll perhaps just write $s\pi$ and $t\pi$ for these.  Question: should I actually conflict with Batanin's usage by using $\delta \pi$ for what I argue it \emph{should} mean, or is this too potentially misleading/confusing?  (Probably, I fear.)  If so, what other notation can I use for the boundary?  $B(\pi)$??  Ugh\ldots

In any case, make sure to explain all this usage in the background introducing pasting diagrams.

\para In ``fundamental $\omega$-groupoid of a type'', what to call $\MLId[X]$?  $\MLId[X]$ is (I think) the best option in isolation (usefully evocative of polynomial rings), but $\globe{0}$ or $\\T[\yon(0)]$ fit better into bigger picture$\ldots$

\para Give the type-theoretic operads catchy names, or just use the definitions (descriptive but unwieldy), or use slightly abbreviated forms of their definitions, i.e.\ $\End(\globes)$ etc.?  Probably this last.

\para Adjunctions? ---remind myself how to place the adjunction symbol them, in each of the various diagram styles I've used!

\section{Terminologies}

\para Set up and explain in the first section: there are many equivalent presentations of a theory.  Main ones for us: syntactic; comprehension cat; cat with attributes.  For categorical models, have plain, based, and stratified.  Show: stratified.  Stratified: full, co-reflective subcategory of based, but not equivalent.  (Mention: Pitts' accessible = 2-equivalent to stratified, but still not equivalent.)  Having got equivalence, explain: \emph{agnostic} about representation; use $\ThId$ for any of the $\strat$ categories (strategories?? $\smiley$)

\para ``pylon diagrama'': fun but probably unnecessary; we already have ``higher spans'', which does the job fine. (Not to mention both ``telescopes'' and ``ladders'', which I abhor; going by which, others might well dislike ``pylon'' just as much!)

\para What term should we use for what $\globes$ form?  A co-weak $\omega$-category, a co-weak-$\omega$-category, a weak co-$\omega$-category, a weak $\omega$-cocategory?  A co-$P$-algebra, or a $P$-co-algebra?

\para Should I conscientiously talk about ``co-operations'' in ``co-endomorphism operads'', or is it clearer to drop the co-?

\section{Checklist}  Notations etc.\ I should check for consistencty during proofing:
\begin{itemize}
\item Any of the notations/terminologies above that I've debated!
\item ``pylon'' $\longrightarrow$ ``(higher) span''
\item \texttt{\\colon} vs.\ :
\end{itemize}

\section{Other}

\para Type-theoretic rules: give in ``hypothetical'' forms.  What does this mean for axioms with no hypotheses?  Shold they be given in the empty context, or in arbitrary context? 
\end{document}