This thesis would not have been possible without the support and encouragement of many people over the past few years.

The Pure and Applied Logic Group at Carnegie Mellon, straddling three departments, provided a marvellously stimulating and diverse intellectual environment within which I was able to gradually find my way---enthusiastically but quite unpremeditatedly---to categorical logic and type theory.  The teaching, advice, and inspiration of James Cummings and Bob Harper were especially helpful.  The unflagging institutional support of the Mathematics department (even when I decided to take an advisor in Philosophy) is also greatly appreciated.

I have also benefited enormously from innumerable discussions with, and support from, colleagues further afield---too many to list, but particularly Martin Hyland, Peter Johnstone, and the other Cambridge category theorists; Thorsten Altenkirch and the type theorists in Nottingham; Mike Shulman and others in Chicago; the Atlantic Category Theory Group at Dalhousie; and, elsewhere, Benno van den Berg and Pierre-Louis Curien.

On a personal level, I would not have made it through a long summer of writing-up without the support of my parents Nicola and David, and a few wonderfully supportive friends---they know who they are!---not to mention UCPO, AUO, the ESO, and the KCO.  To all of them, I am deeply indebted.

Particularly influential on the contents of this dissertation have been Richard Garner, Michael A.~Warren, and Chris Kapulkin, with whom I have greatly enjoyed working over the past few years.

I am grateful also to all the members of my thesis committee for their interest and support, and for bringing a wide range of interests and expertise to bear on this project.  Above all, I would like to thank my advisor, Steve Awodey, for five years of his unstinting support, his guidance, his teaching, generosity, and patience. % \todo{The rhythm of the ending really wants another short line to give a good cadence. Sleep on it!}